---
title: "Final_project-DCW"
author: "David Austin, Collin, Wil Nahra"
date: "2025-04-15"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
``` 

## Introduction

__Diamonds Overview:__ The diamonds dataset focuses on the different qualities of commercial diamonds (carat, color, price, etc.). There are 53940 entries.

## Variables

__price:__ The diamond's price in US dollars

__carat:__ The diamond's weight

__cut:__ The quality of the diamond's cut (Fair, Good, Very Good, Premium, Ideal)

__color:__ The color of the diamond, from D(best) to J(worst)

__clarity:__ The clearness of the diamond (I1(worst), SI2, SI1, VS2, VS1, VVS2, VVS1, IF(best))

__x,y,z:__ The length, width, and depth of the diamond respectively (in mm)

__depth:__ The total depth percentage (z/mean(x,y))

__table:__ The wide of the top of the diamond relative to its widest point

## Specific Questions

__Color and Price:__ The largest question we hope to answer with this dataset is concluding whether the color of a diamond is correlated to its market price. We will answer this by comparing the variables in multiple different ways. Firstly, we show various graphs comparing different variables to price in order to draw a general trend. Then, we will create a box plot comparing color and price to show the spread of data for each color in order to see which color tends to be the most valuable. Then we show a statistics table with data (mean price, count, color, standard deviation price) in order to show empirical proof of our findings. Lastly, we show a p-table to draw whether or not this is caused entirely by chance.

```{r load-data}
library(ggplot2)
data(diamonds)
str(diamonds)
summary(diamonds)
```

## Exploratory Data Analysis

```{r eda-visualizations, message=FALSE, warning=FALSE}
library(dplyr)

# Histogram of Price
ggplot(diamonds, aes(x = price)) + 
  geom_histogram(binwidth = 500, fill = "steelblue", color = "black") +
  ggtitle("Distribution of Diamond Prices")
```

The Histogram above shows the price diamonds from $0 to $20,000 and their popularity. Based on this histogram, we can see visually that the most popular price when it comes to popularity is around the $500 - $1000 range with a steady decrease as the price increases
  
```{r}
# Histogram of Carat
ggplot(diamonds, aes(x = carat)) + 
  geom_histogram(binwidth = 0.1, fill = "salmon", color = "black") +
  ggtitle("Distribution of Carat")
```

This Histogram shows the distribution of carats from 0.1 to 5.0 carats and there popularity/count ranging from 0 to over 10,000. Visualizing the data, we can see when is comes to carat distribution, 0.3 is considered the most popular with a decrease going forward with noticeable burst at the 0.7 and 1.0.

```{r}
# Boxplot of Price by Cut
ggplot(diamonds, aes(x = cut, y = price)) + 
  geom_boxplot(fill = "lightgreen") +
  ggtitle("Price by Cut")
```

The boxplot illustrates the distribution of diamond prices across five cut categories: Fair, Good, Very Good, Premium, and Ideal. The median prices vary noticeably between categories. The "Premium" cut shows the highest median price, followed by "Fair" and "Good," whereas the "Ideal" cut, despite being regarded as the highest quality, has a lower median price. This suggests that while cut quality impacts price, other factors such as carat, color, and clarity also significantly influence the overall value. Additionally, the spread of prices (interquartile range) is similar across cuts, with "Premium" and "Very Good" cuts displaying slightly larger variability.

```{r}
# Scatterplot of Price vs Carat
set.seed(123)
diamonds_sample <- diamonds[sample(nrow(diamonds), 1000), ]

ggplot(diamonds_sample, aes(x = carat, y = price)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  ggtitle("Price vs Carat (Sample of 1000 Diamonds)")
```

This scatterplot visualizes the relationship between diamond price and carat weight for a random sample of 1,000 diamonds from the diamonds dataset. The plot reveals a strong positive linear relationship: as carat increases, price generally rises. While the trend is clear, there’s noticeable spread around the line, especially at certain carat sizes, indicating some variability in pricing even among diamonds of the same weight—likely due to factors like cut, color, and clarity.

```{r}
# Boxplot of Price by Color
ggplot(diamonds, aes(x = color, y = price, fill = color)) +
  geom_boxplot() +
  ggtitle("Price by Diamond Color") +
  theme_minimal()
```

This boxplot visualizes the distribution of diamond prices across different color grades from the diamonds dataset. The x-axis represents distinct diamond colors (D–J), while the y-axis displays price, ranging up to 20,000. Each box encapsulates the interquartile range.
The plot reveals noticeable differences in diamond pricing based on color. Certain color grades exhibit higher median prices, suggesting a potential market preference or rarity effect. The variability in price also differs by color, with some showing a wider range while others are more consistent.

To further investigate this correlation we will use a one-way ANOVA.
Null Hypothesis H_o: The mean price of diamonds is the same across all color grades (D-J).
Alternative Hypothesis H_1: At least one color group has a different mean price.
If the P value is less than 0.05, we will reject the null hypothesis and conclude that color has a significant effect on price.

```{r}
# Statistical Analysis

# Grouped summary statistics by color: mean price, SD(standard deviation), and count
# Helps us see general trends—e.g., which color grades are associated with higher or lower prices
diamonds %>%
  group_by(color) %>%
  summarise(mean_price = mean(price), sd_price = sd(price), count = n())

# Calculate 95% confidence intervals for mean price by color
library(dplyr)

color_ci <- diamonds %>%
  group_by(color) %>%
  summarise(
    mean_price = mean(price),
    sd_price = sd(price),
    n = n(),
    se = sd_price / sqrt(n),
    lower_ci = mean_price - 1.96 * se,
    upper_ci = mean_price + 1.96 * se
  )

# Print the confidence interval table
color_ci

# Plot confidence intervals for mean price by color
ggplot(color_ci, aes(x = color, y = mean_price)) +
  geom_point(size = 3, color = "blue") +
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.3, color = "darkgray") +
  ggtitle("95% Confidence Intervals for Mean Price by Color") +
  ylab("Mean Price (USD)") + theme_minimal()
```

```{r}
# Function for two-sample z-test
z_test <- function(mean1, mean2, sd1, sd2, n1, n2) {
  se_diff <- sqrt((sd1^2 / n1) + (sd2^2 / n2))
  z <- (mean1 - mean2) / se_diff
  p <- 2 * pnorm(-abs(z))  # two-tailed p-value
  return(c(z_score = z, p_value = p))
}

# Compare all unique pairs of colors
color_pairs <- combn(color_ci$color, 2, simplify = FALSE)

# Run z-tests and collect results
z_test_results <- lapply(color_pairs, function(pair) {
  c1 <- filter(color_ci, color == pair[1])
  c2 <- filter(color_ci, color == pair[2])
  
  z_stats <- z_test(c1$mean_price, c2$mean_price, c1$sd_price, c2$sd_price, c1$n, c2$n)
  
  data.frame(
    Comparison = paste(pair[1], "vs", pair[2]),
    Z_Score = round(z_stats["z_score"], 2),
    P_Value = format(z_stats["p_value"], scientific = TRUE, digits = 2),
    Significant = ifelse(z_stats["p_value"] < 0.05, "Yes", "No")
  )
})

# Combine into a data frame and display
z_table <- do.call(rbind, z_test_results)
print(z_table)
```

## Results

__Notable Statistics:__
__95% confidence Intervals (USD)__:  
__D__: (3090.03, 3249.88)  
__E__: (3010.53, 3142.97)  
__F__: (3648.94, 3800.83)  
__G__: (3924.41, 4073.86)  
__H__: (4395.99, 4577.35)  
__I__: (4966.17, 5217.58)  
__J__: (5159.66, 5487.98)  

Using a 95% confidence interval, we observe that diamonds with color grade J have a higher estimated mean price than other color grades, with the interval ranging from $5,159.66 to $5,487.98. We also notice that some color grades, such as D and E, have overlapping confidence intervals, indicating that their mean prices are not significantly different.

__Two sample Z-Test Highlights (z and p scores)__:  
__D vs E__: z = 1.76, p = 7.84e-02  
__D vs F__: z = -9.86, p = 5.90e-23  
__D vs G__: z = -14.85, p = 6.61e-50  
__D vs H__: z = -21.35, p = 3.87e-101  
__D vs I__: z = -25.29, p = 4.29e-141  
__D vs J__: z = -24.06, p = 3.17e-128  
__E vs F__: z = -12.14, p = 1.02e-33  
__E vs G__: z = -17.20, p = 2.81e-66  
__E vs H__: z = -23.16, p = 1.99e-118  
__E vs I__: z = -27.15, p = 1.86e-162  
__E vs J__: z = -26.01, p = 3.64e-149  
__F vs G__: z = -5.06, p = 4.17e-07  
__F vs H__: z = -11.00, p = 1.99e-28  
__F vs I__: z = -15.53, p = 3.57e-54  
__F vs J__: z = -14.51, p = 1.17e-47  
__G vs H__: z = -6.09, p = 1.15e-09  
__G vs I__: z = -10.58, p = 5.26e-26  
__G vs J__: z = -9.43, p = 4.74e-21  
__H vs I__: z = -5.21, p = 1.87e-07  
__H vs J__: z = -4.01, p = 6.07e-05  
__I vs J__: z = -1.38, p = 1.68e-01  

The two-sample Z-tests reveal significant differences in mean diamond prices across most color grade comparisons. Notably, comparisons involving color grades D vs. F through D vs. J all show extremely large negative z-scores (e.g., D vs. J: z = -24.06, p ≈ 3.17e-128), indicating highly significant differences that are far beyond what would be expected by chance.

Similarly, E vs. other lower color grades also display strong significance, with p-values effectively zero. In contrast, comparisons between adjacent high-quality grades, such as D vs. E (z = 1.76, p ≈ 0.078) and I vs. J (z = -1.38, p ≈ 0.168), show no statistically significant difference, as their p-values exceed 0.05.

## Conclusion

__Limitations:__ Put limitations here 

__Potentially Helpful Data Sets:__ Put data sets here


__Potentially Helpful Data Sets:__ Put data sets here
