---
title: "Final_project-DCW"
author: "David Austin, Collin, Will"
date: "2025-04-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
``` 

## Introduction

__diamonds Overview:__ The diamonds dataset focuses on the different qualities of commercial diamonds (carat, color, price, etc.). (Talk more about each variable)

## Specific Questions

__Color and Price:__ The largest question we hope to answer with this dataset is concluding whether the color of a diamond is correlated to its market price. We will answer this by...

```{r load-data}
library(ggplot2)
data(diamonds)
str(diamonds)
summary(diamonds)
```

## Exploratory Data Analysis

```{r eda-visualizations, message=FALSE, warning=FALSE}
library(dplyr)

# Histogram of Price
ggplot(diamonds, aes(x = price)) + 
  geom_histogram(binwidth = 500, fill = "steelblue", color = "black") +
  ggtitle("Distribution of Diamond Prices")

# Histogram of Carat
ggplot(diamonds, aes(x = carat)) + 
  geom_histogram(binwidth = 0.1, fill = "salmon", color = "black") +
  ggtitle("Distribution of Carat")

# Boxplot of Price by Cut
ggplot(diamonds, aes(x = cut, y = price)) + 
  geom_boxplot(fill = "lightgreen") +
  ggtitle("Price by Cut")

# Scatterplot of Price vs Carat
set.seed(123)
diamonds_sample <- diamonds[sample(nrow(diamonds), 1000), ]

ggplot(diamonds_sample, aes(x = carat, y = price)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  ggtitle("Price vs Carat (Sample of 1000 Diamonds)")
# Boxplot of Price by Color
ggplot(diamonds, aes(x = color, y = price, fill = color)) +
  geom_boxplot() +
  ggtitle("Price by Diamond Color") +
  theme_minimal()

# Statistical Analysis

# Grouped summary statistics by color: mean price, SD(standard deviation), and count
# Helps us see general trends—e.g., which color grades are associated with higher or lower prices
diamonds %>%
  group_by(color) %>%
  summarise(mean_price = mean(price), sd_price = sd(price), count = n())

# One-way ANOVA: Tests the null hypothesis that all color groups have the same average price
anova_result <- aov(price ~ color, data = diamonds)
summary(anova_result)
# If the p-value in the summary is < 0.05, we reject the null hypothesis,
# suggesting that at least one color group has a significantly different mean price.
# This means color is likely a relevant factor influencing diamond pricing.

# Post-hoc test (Tukey’s HSD): Performs pairwise comparisons between color groups
TukeyHSD(anova_result)
# If the confidence interval for a color pair does not include 0 and the adjusted p-value is < 0.05,
# then those two color groups differ significantly in mean price.
# This tells us exactly which colors are priced differently from each other.
```

## Results

__Notable Statistics:__ Put it here once we finish the data collection.


## Conclusion

__Limitations:__ Put limitations here 

__Potentially Helpful Data Sets:__ Put data sets here
